```{r cache = T, echo = F, message=F}
# 16Allow knitr to continue to execute even in the presence of errors
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(fig.path = "images/annotating-rayshader-maps-")
```
# Annotating `rayshader` maps

As well as adding computed shading, contour and water overlays to maps, we can also enrich `rayshader` maps with additional annotations such as title information, or, of

## Load in Base Data 

As ever, let's load in our stage data and the elevation raster and create a demo map:

```{r message=FALSE}
geojson_filename = 'montecarlo_2021.geojson'
geojson_sf = sf::st_read(geojson_filename)

library(raster)
library(rayshader)

# Previously downloaded TIF digital elevation model (DEM) file
stage_tif = "stage_elevation.tif"

# Load in the previously saved image raster
elev_img = raster(stage_tif)

# Note we can pass in a file name or a raster object
elmat = raster_to_matrix(stage_tif)

demo_map = elmat %>%
              sphere_shade(texture = "desert")

```

## Adding a Title to a Map

We can add a title to the image as a layer using the `rayimage::add_title()` function [[docs](https://www.rayimage.dev/reference/add_title.html)].

Currently, the title is overlaid on top of the image, so we would need to add a buffer to account for the area taken up by the label if we didn't want to occlude anything in the image by the title. *(Can we calculate how much vertical padding is required?)*

```{r demo-title}
demo_map %>%
  rayimage::add_title("Stage Map", title_size = 20, 
                      title_bar_color = "red",
                      title_bar_alpha=0.8,
                      title_color="white",
                      title_offset = c(120,12), #offset from top left
                      title_style='italic',
                      #title_position='south', # But doesn't relocate bar?
                      ) %>%
  plot_map()
```

Another way of using this feature might be to set the transparency level and show stage results, or example, overlaid on the stage map whilst also seeing the map below.

*It would be useful if we could set the width and location of the title and title_bar [[issue](https://github.com/tylermorganwall/rayimage/issues/3)].*

The title and title bar can also be added (with similar arguments as to `rayimage::add_title()`) via the `plot_map()` function:

```{r demo-title2}
demo_map %>%
  plot_map(title_text='Stage Map',
           title_size = 20,
           title_bar_color = "red",
           title_bar_alpha=0.8,
           title_color="white",
           title_offset = c(120,12), #offset from top left
           title_style='italic',
           #title_position='south', # But doesn't relocate bar?)
  )
```


## Adding Contours to `rayshader` Stage Maps

We have already seen how we can add contours to a raster image. A contour layer can also be created using the `rayshader::generate_contour_overlay()` function:

```{r}
contour_color = "#ffffff" #"#7d4911"

contours_layer = generate_contour_overlay(elmat, color = contour_color, 
                                       linewidth = 1, 
                                       levels=seq(min(elmat),
                                                  max(elmat),
                                                  by=20))
# Increasing the by argument separates out lines;
# This may be important for seeing steep inclines
```

We can then add the overlay in the normal way:

```{r demo-contours}
contoured_map = demo_map %>% 
                  add_overlay(contours_layer,
                              alphalayer = 0.9)

contoured_map %>%  
   plot_map()
```


## Overlaying Image Tiles

As well as creating out own shaded terrain over the raster, we can also overlay tiles retrieved from a 2D map tile server:

```{r demo-overlay, message=FALSE, warning=FALSE}
overlay_image_terrain <-
  geoviz::slippy_overlay(elev_img, 
                 image_source = "stamen",
                 image_type = "terrain",
                 png_opacity = 0.9)

demo_map %>% 
  add_overlay(overlay_image_terrain) %>%
  plot_map()
```


## Plotting Route Layers on `rayshader` Maps

As well as previewing the OSM retrieved roads within the bounding box area of a stage route and the stage routes themselves using simple a `ggplot2` view, we can also render roads and stage routes as layers on a `rayshader` map.

Let's start by creating a line overlay containing the roads in the rendered area:

```{r message=FALSE}
library(osmdata)

stage_osm  = opq(sf::st_bbox(geojson_sf[1,])) %>% 
  add_osm_feature("highway") %>% 
  osmdata_sf()

stage_lines = stage_osm$osm_lines

stage_roads = stage_lines[stage_lines$highway %in% c("unclassified", "secondary", "tertiary", "residential", "service"), ] 


roads_layer = generate_line_overlay(stage_roads,
                                    extent = extent(elev_img),
                                    color="yellow",
                                    heightmap = elmat)
```

We can also create a layer that identifies the stage route, colouring it and setting the line width as required to make it stand out in the rendered view.

In the following example, we create a red coloured stage route layer:

```{r}
red_route = generate_line_overlay(geojson_sf[1,],
                                    extent = extent(elev_img),
                                    heightmap = elmat,
                                    linewidth = 5, color="red")
```

We can now add the roads and stage overlays to our map:

```{r demo-route}
demo_map %>%
  add_overlay(roads_layer)  %>%
  add_overlay(red_route) %>%
  plot_map()
```

This may not look so interesting in a 2D view, but as you will seem it does provide a way of bringing the tile to life when we look at in as an interactive 3D view of 3D video view...
