```{r cache = T, echo = F, message=F}
# 9Allow 1 knitr to continue to execute even in the presence of errors
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(fig.path = "images/route-metrics-")
```

# (PART) Stage Route Metrics {-}

# Stage Route Analysis

This chapter introduces several ideas relating to *stage route analysis* and the generation of "stage route profiles" that describe various features of the the route, such as its twistiness. 

To a certain extent, stage route analysis could be considered as a static analysis complement to the pace notes created for a stage:

```{r}
library("vembedr")
# Hayden Paddon describes pace notes
embed_url("https://www.youtube.com/watch?v=9q7EYVJL9XU")

# See also this Team O'Neil tutorial:
# "Rally Stage Notes and Pace Notes Explained" https://www.youtube.com/watch?v=earjqfXGYzw
```

Stage route analysis also includes a consideration of elevation along the route. Analysis and review of route elevation is explored in more detail in a separate chapter.

## Load in Base Data 

As ever, let's load in our stage data and the elevation raster and create a demo map:

```{r message=FALSE}
library(sf)
library(raster)
library(rayshader)

geojson_filename = 'montecarlo_2021.geojson'
geojson_sf = sf::st_read(geojson_filename)

stage_route_gj = geojsonio::geojson_json(geojson_sf[1,]$geometry)

# Previously downloaded TIF digital elevation model (DEM) file
stage_tif = "stage_elevation.tif"

# Load in the previously saved image raster
elev_img = raster(stage_tif)

# Note we can pass in a file name or a raster object
elmat = raster_to_matrix(stage_tif)

demo_map = elmat %>%
              sphere_shade(texture = "desert",
                           progbar = FALSE)
```

Let's also get a UTM projection of the stage routes:

```{r message=FALSE, warning=FALSE}
# Detect the UTM zone as an EPSG code
lonlat2UTMzone = function(lonlat) {
  utm = (floor((lonlat[1] + 180) / 6) %% 60) + 1
  if(lonlat[2] > 0) {
    utm + 32600
  } else{
    utm + 32700
  }
}

get_utm_projection = function(routes){
  # Keep track of the original proj4 string
  old_crs = st_crs(geojson_sf[1,])$proj4string

  sample_location_x = st_coordinates(st_centroid(routes[1,]))[1]
  sample_location_y = st_coordinates(st_centroid(routes[1,]))[2]
  
  # Generate a new projection in the appropriate UTM zone
  crs_zone = lonlat2UTMzone(c(sample_location_x,
                              sample_location_y))

  new_proj4_string = st_crs(crs_zone)$proj4string

  # Transform the route to the UTM projection
  utm_routes = st_transform(geojson_sf, crs=new_proj4_string)
  
  utm_routes
  # Or should we returned a named list
  # e.g. including the original projection?
  #list(utm_routes = utm_routes, orig_crs=old_crs)
}

utm_routes = get_utm_projection(geojson_sf)
```

In some cases, we may want to discretise a route to smaller segment lengths. We can use the `sf::st_segmentize()` function cut a route described by a linestring into shorter connected segments. For example, we can discretise the route into segments lengths of just 1m long:

```{r}
segmented_route_utm = st_segmentize(utm_routes[1,], units::set_units(1, m))
segmented_route_utm_coords = st_coordinates(segmented_route_utm)
```

Note that if we plot a discretised path, we may end up cutting corners that are described in the original route. Another side effect of discretisation is that the length of the discretised route may change.

## Generating Stage Route Profiles

Although there are no `R` packages (yet!) that are specifically focused on representing and analysing rally stage data, there are several packages designed to support the representation and analysis of animal tracks and route trajectories. These packages are capable of loading and analysing route co-ordinate data specified in a variety of ways and have no difficulty in working with our stage route data.

As to the question of "why bother?", several possibilities come to mind.

In the first case, for the rally fan or rally journalist, learning a bit more about the make up of a stage may add colour when watching or reporting on a stage, or help make sense of why time appears capable of being made up, or lost, on particular stages or particular parts of a stage.

In the second case, an understanding of the detail of a particular stage may help understand strategy, or tyre selection.

In the third case, understanding stage routes may help in setting strategy, or even in planning a stage or hill climb route. With the potential for electric vehicles being used in stage rallies and for hill climbs, energy management and budgeting becomes a key issue. Ensuring that an electric vehicle has enough stored energy to allow it to complete even a single stage, let alone multiple stages with road sections in between, requires an understanding of the likely energy consumption requirements. Knowing the distances involved is one component, but elevation changes may also feed into to energy question, particularly on a a hill climb where the only way is up, and opportunities for regenerative breaking are reduced. Knowing how tortuous a stage is, and the relative mix of fast straights flat corners and slower corners may also help produce simple kinetic energy requirement models for a stage.

### An Aside - Twists and Turns of Rally Stages

Whilst an understanding of the state of the road surface and the current weather conditions may be the key factors in determining tyre choice, how "twisty and turny" a stage route may also influence tyre selection and driving style. 

The tread pattern of rally tyres are designed to support acceleration even when turning, so if tyres are wearing thin and the last stage of day has far more aggressive right corners than turns to a left, knowing the stage profile may help save a second or two.

```{r tyre-video, message=FALSE}
embed_url("https://www.youtube.com/watch?v=-I74uopkcro")
```

## Finding Stage Route Lengths

To begin with, let's start with a very simple metric, *stage route length*.

We can find the length of each stage route in meters using the `sf::st_length()` function:

```{r}
st_length(geojson_sf)
```

Or for a single stage:

```{r}
st_length(geojson_sf[1,])
# 21069.62
```

The `rgeos::gLength()` function will generate the length of a *Spatial* object using its natural units. Which is to say, we'd need to project into something like UTM co-ordinates first if we wanted to use it to calculate the distance in meters.

We can also find the route length directly from a geojson string that encodes latitude and longitude locations using the *turf.js* like `geoops::geo_line_distance()` function:

```{r}
route_as_geojson = geojsonio::geojson_json(geojson_sf[1,]$geometry)

geoops::geo_line_distance(route_as_geojson)
# 21.05409
```


The `geosphere::distGeo()` function provides a way of calculating distances for each step of the route represented using the lonlat projection:

```{r}
# Get the coordinates for each step
lonlat_coords = st_coordinates(geojson_sf[1,])[,1:2]

# Omit the final NA value
route_step_lengths = head(geosphere::distGeo(lonlat_coords), -1)

# Calculate the accumulated distance at each step
# and return the final (maximum) value:
max(cumsum(route_step_lengths))
# 21069.62
```

One thing we notice is that some of these distances are: a) different to each other; b) different to the officially stated stage distance of 20.58 km.

The projection and parameters used in a projection to calculate the distance all contribute to the actual calculated value. In addition, our distance calculations are base on summing the lengths of the multiple straight line segments described by the route data rather than the tighter (shorter) racing line. It's also not clear from the route data where the extent of the route is the extent of the competitive portion of the stage, or whether the start and end points refer to the time control locations rather than competitive timing lines.

On the other hand, the  actual route is likely to be longer than what we might term the "ground distance" as you can tell from an elevation chart that shows elevation of the route against distance into the route. If we elevation and distance axes on the same scale, the elevation line is unlikely to be a straight line: the elevation may go up or down. If we straighten out the elevation line to a flat line and place on the x-axis, it will be longer than the original "on the flat" route distance.


## Using Ecological Movement Analysis Tools to Analyse Stage Routes

Ecologists researching animal movements have produced a wide range of tools for categorising and classifying the trajectories of routes followed by one or more animals. Trajectories are typically recorded as sequences of locations recorded at particular times.

The [`adehabitatLT` *R* package](https://cran.r-project.org/web/packages/adehabitatLT/index.html)  usefully categorises animal trajectories in the following ways:

- *type I*: location data is available but a sample timestamp is not precisely known or not available for each step;
- *type II*: as well as location data, a timestamp for each step is also available; this type is then further subdivided into:
  - *regular trajectories*, where there is a constant delay or period between each step sample;
– *irregular trajectories*, where there is a variable delay between each step sample.

As far as rally data goes, stage route data may be trivially classed as either *type 1*. However, if we use distance into route as a time proxy, we can also create  *type II irregular* traces; telemetry data should ideally be *type II regular*.

One intuition behind using animal movement analysis tools to categorise rally stage routes is that we can analyse the route as if it were a *type 1* style animal track. Another intuition is that we can use distance into route as a dummy time variable (e.g. use a dummy equivalence mapping of 1m into route as 1s, although we could also attempt to use proxy times, such as setting *time = distance_into_stage x nominal_speed*, or even use a speed model that may depend on route properties at each step of the stage).

## Using `trajr` for Stage Route Profiling

The [`trajr` *R* package](https://cran.r-project.org/web/packages/trajr/vignettes/trajr-vignette.html) was developed to help characterise and analyse the trajectories or paths followed by a moving animal given various location coordinates along its route.

Trajectories are created using Cartesian co-ordinates, so it makes sense to use data that has been cast into UTM coordinate system as the basis for the trajectory. A two dimensional spatial `trajr` trajectory can be created by default from a two column (x/y coordinates) or three column (x/t/time) dataframe using the `trajr::TrajFromCoords()` function, r from an arbitrary dataframe by specifying column names using the *xCol*, *yCol* and *timeCol* parameters. Spatial and temporal units can also be specified using the *spatialUnits* and *timeUnits* parameters.

```{r}
library(trajr)

trj <- TrajFromCoords(as.data.frame(st_coordinates(utm_routes[1,])))
```

To cast all the routes to trajectories, we could do something like the following:

```{r}
get_trj = function(route){
  TrajFromCoords(as.data.frame(st_coordinates(route$geometry)))
}

trjs = apply(utm_routes, 1, get_trj)
```

We can retrieve the stage length using the `trajr::TrajLength` function:

```{r}
TrajLength(trjs[[1]])
```

We can create also create a simple plot of the trajectory, and overplot it with the location points used to construct the route:

```{r trajr-plot}
# Create a simple plot the route
plot(trj)

# Overlay points onto the plot
points(trj, draw.start.pt = FALSE,
       #pch = 16,  # symbol
       col = "black", 
       cex = 0.4 # scale
       )
```

For each fragment of the line we can calculate the distance from its displacement, or directly using `trajr::TrajStepLengths` function, and then cumulatively sum them to get the overall route distance so far:

```{r}
# displacement is a complex number, so we can get the actual distance:
trj$distance = Mod(trj$displacement)

# Find the distance of the just completed step
trj$distance2 = c(0, TrajStepLengths(trj))

# Find the distance of the upcoming step
trj$predist = c(TrajStepLengths(trj), 0)

# Find the accumulated distance at each step
trj$cum_dist = cumsum(trj$distance)

head(trj[,c('distance', 'distance2', 'predist', 'cum_dist')])
```


### Smoothing a Trajectory

Sometimes a stage route may appear rather "ragged" as an artefact of using multiple straight line segments to construct the route.

We can smooth a route with a Savitzky-Golay smoothing filter called from `signal::sgolayfilt` using the `trajr::TrajSmoothSGorder(order, filter_length_odd)` function.

Using a third order filter applied to a window size of 5 steps retains quite a smooth fitting line whilst removing some of the sharp edges caused by concatenating straight line segments

```{r trajr-plot-smoothed}
smoothed <- TrajSmoothSG(trj, p=3, n=5)
#p polynomial order
#n filter length (window size); must be an odd integer
plot(trj)
plot(smoothed, col = "red", add = TRUE)
```

The [`smoothr` *R* package](https://cran.r-project.org/web/packages/smoothr/vignettes/smoothr.html) also provides several useful smoothing utilities.

So that we can more clearly see what's going on, let's grab just a fragment of our stage route:

```{r}
get_route_segment = function(i, steps){
  cc = as.data.frame(st_coordinates(geojson_sf[i,]))[steps,]

  st_linestring(cbind(x=cc[,1], y=cc[,2]))
}

route_segment = get_route_segment(1, 1:10)
```

For example, the `smoothr::smooth` function allows us to apply various smoothing approaches. Let's see how well its *ksmooth* kernel smoothing approach works, that applies a Gaussian kernel regression to the x and y coordinates independently:

```{r smoothr_ksmooth}
smoothr_route_segment = smoothr::smooth(route_segment,
                                        method = "ksmooth")

plot(st_geometry(route_segment), col = 'grey', lwd = 3)
plot(st_geometry(smoothr_route_segment),
     col = "red", add = TRUE)
```

Alternatively, we could use the `smoothr` *spline* method, which calculates a spline interpolation over the x and y coordinates independently:

```{r smoothr_spline}
plot(st_geometry(geojson_sf[1,]), col = 'grey', lwd = 3)
plot(st_geometry(smoothr::smooth(geojson_sf[1,], method = "spline")),
     col = "red", add = TRUE)

smoothr_route_segment2 = smoothr::smooth(route_segment, method = "spline")

plot(st_geometry(route_segment), col = 'grey', lwd = 3)
plot(st_geometry(smoothr_route_segment),
     col = "red", add = TRUE)
```

For a more detailed look at smoothing trajectories and finding racing lines, albeit in circuit racing, see  *Heilmeier, A., Wischnewski, A., Hermansdorfer, L., Betz, J., Lienkamp, M. and Lohmann, B., 2019. Minimum curvature trajectory planning and control for an autonomous race car. Vehicle System Dynamics. DOI: 10.1080/00423114.2019.1631455* and the related Python laptime simulation package, [`TUMFTM/laptime-simulation`](https://github.com/TUMFTM/laptime-simulation).

### Resampling Locations

We can use the `trajr::TrajRediscretize` function to resample locations from the line that are a fixed distance apart as the crow flies. s well as finding sample points, a new trajectory is also calculated that connects those points.

For example, if we discretise the route to points 500m apart as the crow flies, we note that the newly discretised route now has straight line segments between each of the sample points. This is perhaps not so good for stage route analysis, but it does provide one way of obtaining sample point locations from telemetry data. Ideally, though, sample points would be a fixed distance apart *along the route*:

```{r trajr-rediscretized}
# Resample to step length 500m
resampled <- TrajRediscretize(trj, 500)
plot(trj)

# Plot rediscretized trajectory in red
lines(resampled, col = "#FF0000A0", lwd = 2)
points(resampled, type = 'p', col = "#FF0000A0", pch = 16)
legend("topright", c("Original", "Rediscretized"),
       col = c("black", "red"), 
       lwd = 2, inset = c(0.01, 0.02))
```

Can we see how many splits there are?

```{r}
nrow(resampled)
```

### Resampling Against Distance

The `trajr` package doesn't seem to support resampling by distance along the line but it does allow resampling by time. So if we set the time to the accumulated basis, we can use time as a proxy for distance...

```{r}
# Backup the original (dummy) timestamps
trj$displacementTime_ = trj$time

# Use the accumulated distance as a time proxy
trj$time = trj$cum_dist
```

Let's see what happens if we resample based on this use of time as a proxy for distance sampling the route every 2km:

```{r trackr-resampled}
# The time units for the route are in seconds: TrajGetTimeUnits(trj)
resampled_t = TrajResampleTime(trj, 2000 )
plot(trj)

#lines(resampled_t, col = "#FF0000A0", lwd = 2)
points(resampled_t, type = 'p', col = "#FF0000A0", pch = 16)
legend("topright", c("Original", "Resampled"),
       col = c("black", "red"), 
       lwd = 2, inset = c(0.01, 0.02))
```

### Splitting a Trajectory

We can split a trajectory using the `trajr::TrajSplit()` function. This takes a trajectory and an index of split points along the path and returns a list of trajectories containing the segment starting from each of the index values to the next supplied value.

If we try to use this function to split the route by split points applied generated by discretising or sampling the route, the resolution of the original line is lost: we are not cutting the *original() line at the sample points, we are cutting the *new* line created from the sample points:

```{r split-resampled}
plot(TrajSplit(resampled_t, c(1, 2, 4))[[3]])
```

However, we *can* cut the original lines using the `stplanr::route_split()` function:

```{r split-route-colours}
#https://www.rdocumentation.org/packages/stplanr/versions/0.6.2/topics/route_split
library(stplanr)

# Get the second resampled point
pt = st_sfc(st_point(c(resampled_t[2,'x'],resampled_t[2,'y'])),
             crs=st_crs(utm_routes[1,])$proj4string)

# Get the third resampled point
pt2 = st_sfc(st_point(c(resampled_t[3,'x'],resampled_t[3,'y'])),
             crs=st_crs(utm_routes[1,])$proj4string)

# Plot the original route in grep
plot(utm_routes[1,]$geometry, lwd = 9, col = "grey")

# Split the route about the second point
split_route = route_split(utm_routes[1,], pt)

# And plot the two parts
plot(split_route$geometry, col = c("red", "blue"), add = TRUE)
```

We can view each line segment separately:

```{r segmented-stage-parts}
# Use two columns to display plots side by side
par(mfrow=c(1,2))

plot(split_route[1,]$geometry)
plot(split_route[2,]$geometry)
```

We can automate the extraction of segmented lines using the following approach, which takes a route, splits it by the first point, splits the second half of the segment by the second point, and so on. Note that this requires us to order the points we want to split the line along in the order they appear along the route:

```{r}
split_line = function(route_, pts_) {
  segments = c()
  for(i in 1:length(pts_)) { 
    split_ = route_split(route_, pts_[i])
    segments = c(segments, split_[1])
    route_ = split_[2]
  }
  route_
}

# Points we wish to segment on
pts = c(pt, pt2)

split_route2 = split_line(utm_routes[1,], pts)

split_route2
```

Now we can plot out three segment line (two split points):

```{r three-segment-line}
plot(utm_routes[1,]$geometry, lwd = 9, col = "grey")
plot(split_route2$geometry, col = c("red", "blue", "green"), add = TRUE)
```

### Rotating the Stage Trajectory

To visualise routes in a similar profile, we can use the `trajr::trajr-rotate()` function to align the start and finish points. An *angle* parameter gives the angle in radians between the first and last points.

For example, for a horizontal, left to right right stage view:

```{r trajr-rotate-horiz}
plot(TrajRotate(trj, angle = 0, relative = TRUE))
```

Or for a view where we read the stage up from the bottom:

```{r trajr-rotate-vert}
plot(TrajRotate(trj, angle = pi/2, relative = TRUE))
#Arguments
#trj The trajectory to be rotated.
#angle The angle in radians between the first and last points in th
```
It might be handy to rebase the coordinates of the start to (0,0).

### Straightness and Sinuosity

Several simple measures exist that try to provide a single number measure describing something about the straightness or sinuosity of a trajectory.

For example, the `trajr::TrajStraightness()` function determines *straightness* from the start to the end of a route as the ratio *D/L*, where *D* is the distance from the start to the end of the trajectory, and *L* is the length of the trajectory:

```{r}
TrajStraightness(trj)
```

We can get the straightness for each stage as follows:

```{r}
utm_routes['straightness'] = unlist(lapply(trjs, TrajStraightness))
utm_routes[,c('Name', 'straightness')]
```

This suggests that SS 5 is perhaps the most least straight, and SS13/15 is most straight.

If reading data from tables is not your thing, here's a chart...

```{r straight_bar, message=FALSE}
library(ggplot2)

g1 = ggplot(utm_routes) +
        geom_col(aes(x=as.numeric(row.names(utm_routes)),
                     y=straightness)) +
  scale_x_discrete(limits=utm_routes$Name) + xlab('Stage')

g1
#ggpubr::ggarrange(c(g1, g2))
```

Let's render them both, with a simple indicator to show where the *start* of each stage is.

```{r straightness-stages, warning=FALSE, message=FALSE}
start_location = function(i) {
  as.data.frame(st_coordinates(utm_routes[i, ]$geometry))[1,]
}

plot_stage = function(i, label){
   ggplot(data=utm_routes[i,]) + 
          geom_sf() + 
          geom_point(aes(x=X, y=Y), size=2,
                     col='red', alpha = 0.5,
             data=start_location(i)) +
          ggtitle(label) + coord_sf()
}

g_SS5 = plot_stage(5, 'SS5 - least straight' )

g_SS13 = plot_stage(9, 'SS13/15 - most straight')

ggpubr::ggarrange(g_SS5, g_SS13)
```

Note that where we have multiple split point tines and locations for a stage available, we could slice the stage into split segments and analyse the straightness of each one separately.

Another single value measure is the *sinuosity*, "a function of the mean cosine of the turning angle", calculated in a corrected form that does not require constant step size by the [`traj::TrajSinuosity2()` function](https://rdrr.io/github/JimMcL/trajr/man/TrajSinuosity2.html) (based on equation 8 of *Benhamou, S. (2004). [How to reliably estimate the tortuosity of an animal's pat](http://www.seaturtle.org/library/BenhamouS_2004_JTheorBiol.pdf), Journal of Theoretical Biology, 229(2), 209-220. doi:10.1016/j.jtbi.2004.03.016*):

Let's see how the sinuosity compares with straightness on each stage:

```{r}
utm_routes['sinuosity'] = unlist(lapply(trjs, TrajSinuosity2))
utm_routes[, c('Name', 'straightness', 'sinuosity')]
```

Again, if reading data from tables is not your thing, we can visualise how straightness and sinuosity relate on each stage using a scatterplot:

```{r str_sin_scatter}
g2 = ggplot(utm_routes, aes(x=straightness,
                            y=sinuosity,
                            label = Name)) + 
        geom_point(size=1)  +
        # hjust: left justification (0), right justification (1)
        geom_text(size=3, hjust = 1, nudge_y = -0.001)
g2
```

Under this measure we see that the most sinuous state is actually SS 13/15 and SS5 also has high sinuosity, whereas SS 2 is the least sinuous stage:

```{r trajr_sinuosity}
g_SS2 = plot_stage(2, 'SS2 - least sinuous')

g_SS13b = g_SS13 + ggtitle('SS13 - most sinuous')

ggpubr::ggarrange(g_SS2, g_SS13b) 
```

### Directional Change

Another possible measure of how twisty a route is the *directional change*. As this measure is defined for each pair of steps, the directional change measure applied to a route is given as the mean (DC) and standard deviation (SDDC) calculated over all directional changes. Directional change itself is defined as the angular change (in degrees) between any two points in the trajectory, divided by the time difference between the two points. According to the [`trajr` vignette](https://cran.r-project.org/web/packages/trajr/vignettes/trajr-vignette.html), *DC may be used as an index of nonlinearity, and SDDC as a measure of irregularity*.

The time basis of this measure means we have to use some sort of poetic license when using this measure for route analysis: do steps go up one time unit per step, or do we use the accumulated distance for time stamps?

```{r}
calc_dc = function(route){
  mean(TrajDirectionalChange(route))
}

calc_sd = function(route){
  sd(TrajDirectionalChange(route))
}

utm_routes['DC'] = unlist(lapply(trjs, calc_dc))
utm_routes['SDDC'] = unlist(lapply(trjs, calc_sd))
utm_routes[, c('Name', 'DC', 'SDDC')]
```

Exploring these measures with different time bases is left for another day...

### Tracking Turning Angles
One final single value, albeit, a complex value, metric we can get is the mean vector of the turning angle, given via the `trajr::TrajMeanVectorOfTurningAngle()` function.

This function takes a *compass.direction* (default `NULL`) which indicates whether the angle should be calculated relative to the angle (in radians) or, if `NULL`, relative to the previous step.

```{r}
utm_routes['meanTurn'] = unlist(lapply(trjs,
                                       TrajMeanVectorOfTurningAngles))

utm_routes[, c('Name', 'meanTurn')]
```

For example, SS2 and SS13 appear to rotate in different directions?

```{r trajr-mean-turn}
ggpubr::ggarrange(g_SS2, g_SS13) 
```

Hmm... maybe!

We can also take the mean vector and turn it into polar coordinates. The angle is then the mean angle of turn (`Arg(meanTurn)`), although the magnitude (`Mod(meanTurn)`) for the relative angle summary is perhaps less useful.

Parsing the angle in terms of the general direction of travel over the course of the route provides a potential signal regarding the general compass direction of travel. This may be relevant when trying to identify whether a driver might be driving into the sun, for example (recall that we can look up sun angle and elevation by date and time easily enough using the `suncalc` *R* package.

The simplest calculation is probably just to calculate the bearing between the start and the end point of the route, for example using the `geosphere::bearing()` function.

Alternatively we might consider a summary statistic of the amount of turn at each step along the route.

For example, we can calculate the angle of each step on a route relative to a specific compass angle, which I think(?!) has the compass angle of 0 radians for due North:

```{r}
compass_relative_turn = function(route, angle=0) {
  TrajMeanVectorOfTurningAngles(route, angle)
}

utm_routes['meanCompass'] = unlist(lapply(trjs,
                                          compass_relative_turn))

utm_routes[, c('Name', 'meanCompass')]
```

In this case the angle will give the average compass direction of travel and the modulus may give us an indication of the straightness of travel in that direction?

So... do SS2 and SS6 appear to go in different directions?

```{r trajr-mean-compass-different}
g_SS2 = plot_stage(2, 'SS2 - one way..')

g_SS5 = plot_stage(5, 'SS5 - and another?')

ggpubr::ggarrange(g_SS2, g_SS5) 
```

And do SS9 and SS11 appear to go in roughly the same direction?

```{r trajr-mean-compass-same}
g_SS9 = plot_stage(6, 'SS9 - and the same?')

ggpubr::ggarrange(g_SS2, g_SS9) 
```

### Visualising Trajectory Direction Changes

As well as providing single number metrics, we can also get information back at the level of each step of the trajectory. We have already seen how the complex *displacement* is given for each step, from which we can calculate the distance of each step.

We can also get access to the angle of each step in a route using the `trajr::TrajAngles()` function. This function also takes a *compass.direction*, again indicating whether the angle should be calculated relative to the angle (in radians) or, if `NULL`, relative to the previous step.

```{r}
# Step angle in radians relative to previous
trj$stepangle = c(0, TrajAngles(trj, compass.direction = NULL) * 180 / pi, NA) 

trj$cumstepangle = cumsum(c(0, TrajAngles(trj, compass.direction = NULL) * 180 / pi, NA))
trj$stepheading = c(TrajAngles(trj, compass.direction = 0)* 180 / pi, NA) 


head(trj[,c('stepangle', 'cumstepangle', 'stepheading')], 5)
```

We can plot the route using colour to give the general heading, by quadrant:

```{r trajr-route-headings, warning=FALSE}
# Cut styling via https://stackoverflow.com/a/29966785/454773

ggplot(data=tidyr::drop_na(trj, stepheading ),
       aes(x=x, y=y, color=col)) + 
  geom_point(aes(colour = cut(stepheading,
                              c(-180, -90, 0, 90, 180))),
             size = 1) +
  coord_sf() + 
  geom_point(aes(x=X, y=Y), size=1, col='black',
             data=start_location(1))
```

Can we identify tight turns? The following chart plots each step location, using a distinct marker to highlight turns with a large *stepangle*:

```{r trajr-tight-turns, warning=FALSE}
ggplot(data=trj,
       aes(x=x, y=y), size=0.5) + geom_path(color='grey')+
  geom_point( pch=0, color='blue',
             data=trj[abs(trj$stepangle)>45,]) +
    geom_point(data=trj[abs(trj$stepangle)<=45,],
                pch=1, color='red', size=0.1) +
  coord_sf() + 
  geom_point(aes(x=X, y=Y), size=1, col='black',
             data=start_location(1))
```

The instantaneous angle is okay for trying to identify particularly tight turns, but to identify whether a corner increase the angle in the same step direction, which is to say, whether the corner *tightens*, for example, we need process the data a bit further.

One way of doing this is to determine the gradient of the accumulated angle using the [`pracma::gradient` function](https://www.rdocumentation.org/packages/pracma/versions/1.9.9/topics/gradient). 

```{r}
trj$step_gradient = pracma::gradient(trj$cumstepangle, trj$cum_dist)
```

The sign of the gradient then identifies the turn direction but the magnitude of the gradient is harder to parse because the step sizes are not constant. 

A plot using colour to depict gradient sign may help identify whether a route involves turn predominantly to the left or to the right. We could use a related approach to display a table of locations labeled as tight right or left corners, for example:

```{r trajr-gradient-lr, warning=FALSE}
ggplot(data=trj,
       aes(x=x, y=y)) + geom_path(color='grey')+
  geom_point( size=0.2, color='blue',
             data=trj[trj$step_gradient>0.2,]) +
    geom_point(size=0.2,
               data=trj[trj$step_gradient<=-0.2,],
               color='red') +
  coord_sf() + 
  geom_point(aes(x=X, y=Y), size=1, col='black',
             data=start_location(1))

```

If nothing else, however, this approach also gives us an immediate way of detecting stationary points in the gradient. As a consequence, we can identify directional changes from left to right, or *vice versa*.

```{r trajr-gradient, warning=FALSE}
ggplot(trj,aes(x=cum_dist)) + 
    geom_hline(yintercept=0, color='grey' ) +
    geom_hline(yintercept=-500, color='grey' ) +
    geom_line(aes( y=cumstepangle)) +
    geom_line(aes( y=-500+50 * step_gradient),
              color='red')
```

By creating simple rules, we might be able to identify features like *tightens* (for example, if we have three angles in the same direction with increasing magnitude and the final angle exceeding some threshold limit). We might similarly be able to create rules around elevation data, for example to identify *crests* (for example, a maximum in elevation, found by finding stationary points of elevation / first derivative).

Finally, it's worth briefly mentioning a couple more `trajr` functions that may be useful when it comes to analysing telemetry data: `trajr::TrajVelocity` and `trajr:TrajAcceleration`, which estimate velocity and acceleration as vectors at each point along a trajectory. But that's for another book...

## Using `amt` for Stage Route Profiling

The [`amt` (animal movement tools) *R* package](https://cran.r-project.org/web/packages/amt/index.html) was designed to support the management and analysis of animal movement data. Routes are represented as *tracks* in the form of tibbles with at least an *x_* and *y_* column and, if required, a `t_` (time) column. Consecutive rows, which might be assumed to be collected with fixed period, identify straightline *steps*. A `track_xy`track type only has coordinates, and a
`track_xyt` track type has a timestamp associated to each coordinate pair.

An additional *burst* column, `burst_`, provides us with a potential grouping parameter that lets us perform analyses on sections of the track. This could be used to identify segments of the route between split points, for example, if we can apply appropriate burst labels to each part of the track.

For our purposes here, it is clear to see how we might imagine a mapping the X and Y co-ordinates from a route linestring in a UTM/meters based projection directly into a a track tibble. The `amt::make_track()` function provides a straightforward way of doing this:

```{r}
library(amt)

amt_track = make_track(st_coordinates(utm_routes[1,]), X, Y)
amt_track
```

### Creating Tracks

The `make_track` function has a further optional argument (`crs`), which allows the user to set a coordinate reference system (CRS) of the track. The CRS needs to be provided as valid proj4string, see the documentation of sp::CRS for further details; see also [*proj4* projections](https://proj.org/usage/projections.html).

We can segment the track into a series of straightline steps from one point to the next, also returning the step distance (`sl`) and angle (`direction_p` radians):

```{r}
# Create a "steps" object from each sample
amt_steps = steps(amt_track)
head(amt_steps)
```

### Converting to Other Track Formats

The `amt` package provides a range of tools that allow us to *coerce* `amt` tracks to other movement analysis packages.

For example, we can cast a `track_xy` object to a *Spatial* object:

```{r}
as_sp(amt_track)
```

Conversions to other movement data structures is possible from `track_xyt` objects.

## Using `amt` for Stage Route Analysis
We can also obtain step lengths directly from the track via the `amt::step_lengths()` function:

```{r}
# Obtain the step lengths and omit / drop the final NA value
amt_step_lengths = head(step_lengths(amt_track), -1)
```

Directions can be determined absolutely, e.g. with reference to a North, South, East or West basis, using `amt::direction_abs()` or relatively to the previous step using ``amt::direction_rel()`:

```{r}
absolute_step_directions = direction_abs(amt_track)
relative_step_directions = direction_rel(amt_track)
```

In terms of route complexity measures, the `amt` package provides functions for calculating single number metrics of *sinuosity* (`amt::sinuosity()`), 
*straightness* (`amt::straightness()`), *accumulated track distance* (`amt::cum_dist()`) and *total straightline distance* (`amt::tot_dist()`). The *intensity use* metric is calculated by dividing the total movement distance (`tot_dist`) by the square of the area of movement:

```{r}
amt::sinuosity(amt_track)
amt::straightness(amt_track)
amt::cum_dist(amt_track)
amt::tot_dist(amt_track)
amt::intensity_use(amt_track)
```

Let's generate a list of tracks, one for each stage:

```{r warning=FALSE, message=FALSE}
to_amt_track = function(route){
  make_track(st_coordinates(route$geometry), X, Y)
}

amt_tracks = apply(utm_routes, 1, to_amt_track)
```

Now we can generate reports for all the stages:

```{r}
utm_routes['amt_sin'] = unlist(lapply(amt_tracks, amt::sinuosity))
utm_routes['amt_str'] = unlist(lapply(amt_tracks, amt::straightness))
utm_routes['amt_cumd'] = unlist(lapply(amt_tracks, amt::cum_dist))
utm_routes['amt_totd'] = unlist(lapply(amt_tracks, amt::tot_dist))
utm_routes['amt_int'] = unlist(lapply(amt_tracks, amt::intensity_use))

utm_routes[, c('amt_sin','amt_str','amt_cumd','amt_totd','amt_int')]
```

We can resample locations from the track using `amt::track_resample()`. Once again, if we use distance as a proxy for time, this allows us to sample by distance along the line.

## Using `rLFT` for Stage Route Profiling

The [`rLFT` (Linear Feature Tools) *R* package](https://cran.r-project.org/web//packages/rLFT/vignettes/rLFT_Introduction.html) was developed in order to support the processing of simple linear features.

The boundary convexity tool function, `rLFT::bct()` generates a convexity measure for a sliding window of points along a route; the route must be provided in a projected (*not* geographic) coordinate system. The sign of the convexity measure identifies whether the curve is concave (negative value, left turn) or convex (positive value, right turn), the increasing values identifying tightness of the curve.  A normalised convexity index and a sinuosity measure  are also created.

The `bct()` function takes two important arguments:

- the *step*, which describes the distance between measurements along the route (i.e. the distance between ample points), and
- the *window*, which describes the size of the window (the distance along the route) used to measure the convexity value.

```{r rlft-route}
library(rLFT)

route_convexity <- bct(utm_routes[1,],
                   # distance between measurements 
                   step = 10,
                   window = 20, ridName = "Name")

head(route_convexity)
```

We can get the output table as a *simple features* object where the midpoint co-ordinates are cast to a point geometry type using the `sf:st_as_sf()` function:

```{r}
outSF <- st_as_sf(route_convexity,
                 coords = c("Midpoint_X", "Midpoint_Y"), 
                 stringsAsFactors = FALSE)
```

We note that by sampling the actual route at 10m intervals when calculating curvature *and then recreating a route from those 10m sample points*, we may actual crop the end of tight corners as represented by the original route path. Reducing the step size would provide a more faithful route (recall, the *window* size is actual used in performing the convexity calculation). Arguably, it might make sense to use generate a smoothed route from the original route and then use a small step size over the smoothed route as the basis for all route based metrics.

The convexity measures are defined in [*Measuring boundary convexity at multiple spatial scales using a linear 'moving window' analysis: an application to coastal river otter habitat selection*](https://www.d.umn.edu/biology/documents/Nibbelink2_000.pdf), Albeke, S.E. et al., Landscape Ecology 25 (2010): 1575-1587.


### Convexity Profile of a Stage Route

We can profile the convexity of a stage route in the terms of the convexity against the distance along the route in terms of a simple line chart. Recall that the sign of the convexity value denotes the direction of a corner (negative left, positive right).

We can add a grey bar with low alpha transparency to slightly mask the less sever corners:

```{r curvature-profile-bar}
ggplot(route_convexity) +
  geom_bar(aes(x=MidMeas, y=ConvexityIndex, col=(ConvexityIndex>0)),
           stat='identity', show.legend = FALSE) +
  geom_rect(ymin = -0.1, ymax = 0.1, 
            xmin = -Inf, xmax = Inf,
            fill = 'lightgrey', alpha=0.05)
```

### Previewing Significant Corners

By filtering the table to rows with with a large absolute *ConvexityIndex* we can identify and label tight corners. The `ggrepel::geom_text_repel()` function attempts to minimise the extent to which labels overlap, using arrows to identify points referenced by the label where the label location is dodged away from the point:

```{r rLFT-route, message=FALSE, warning=FALSE}
tight_corners = route_convexity[abs(route_convexity$ConvexityIndex)>0.45,]

ggplot() +
  geom_sf(data=utm_routes[1,]) +
  ggrepel::geom_text_repel(data=tight_corners,
                           aes(label = ConvexityIndex,
                               x=Midpoint_X, y=Midpoint_Y,
                               color= (ConvexityIndex>0) ),
                           size=2,
                           nudge_x = 2000, nudge_y = 500) +
  geom_point(data=tight_corners,
             aes(x=Midpoint_X, y=Midpoint_Y,
                 color= (ConvexityIndex>0) ), size=1) +
  theme_classic()
```

### Convexity as a 3D Plot

To try to get a feel for how the convexity changes along a route, we can render a 3D plot of the of the convexity using the `rayhader::ggplot()` function.

As a first step, we need to create a base *ggplot2* plot:

```{r convexity_gg}
convexity_gg = ggplot() + geom_sf(data=utm_routes[1,]) +
  geom_point(data=route_convexity,  show.legend = FALSE,
             size=1,
            aes(x=Midpoint_X, y=Midpoint_Y,
                color=abs(ConvexityIndex)))

convexity_gg
```

Then we can render it in three dimensions:

```{r}
library(rayshader)
library(rgl)

rgl::clear3d()
plot_gg(convexity_gg, scale=1000, raytrace = FALSE)

rgl::rglwidget()
```

Looking at these metrics over a complete route can be a little hard to decipher. However, if we segment the route into smaller sections, as we shall see in the next chapter, we will see how can start to use these metrics to describe the route in some useful detail.

## Finding Curvature of a Route

The `pracma::circlefit()` function will attempt to calculate the center and radius of a curvature from an arc defined by a path segment described by three or more consecutive locations provided as *x* and *y* coordinate lists. We can use the radius of curvature to determine corner severity numbers that are typically used in rally pace notes.

For example, we can build on the the `rLFT` segmented route to add the radius of curvature of turns starting at each midpoint:


```{r results=FALSE}
# The curvature function takes an arc defined over
# x and y coordinate lists
curvature = function(x,y){
  #729181.8, 729186.1, 729190.4
  #4957667 , 4957676, 4957685
  tryCatch({
      # circlefit gives an error if we pass a straight line
      # Also hide the print statement in circlefit
      # circlefit() returns the x and y coords of the circle center
      # as well as the radius of curvature
      # We could then also calculate the angle and arc length
      pracma::circlefit(x,y)[3]
    },
    error = function(err) { 
      # For a straight, return the first co-ord and Inf diameter
      # Alternatively, pass zero diameter?
      c(x[1], y[1], Inf)[3]})
}

curvature2 = function(x1, x2, x3, y1, y2, y3){
  curvature(c(x1, x2, x3), c(y1, y2, y3))
}

# The base::Vectorize function provides a lazy way of 
# vectorising a non-vectorised function
curvatures = Vectorize(curvature2)

# Tests
#curvature(c(729181.8,729186.1,729195.4),c(4957667,4957676,4957685))
#curvature(c(729194.6,729198.8,729199.6),c(4957694,4957703,4957713))
```

Let's find the radius of curvature for each 10m sampled point we obtained when finding convexity values along the route:

```{r message=FALSE, warning=FALSE, results=FALSE}
library(dplyr)
library(pracma)
  
route_convexity$radius = curvatures(lag(route_convexity$Midpoint_X), 
                                    route_convexity$Midpoint_X,
                                    lead(route_convexity$Midpoint_X),
                                    lag(route_convexity$Midpoint_Y),
                                    route_convexity$Midpoint_Y,
                                    lead(route_convexity$Midpoint_Y)
                                    )
```

Rally pace note systems tend to use a simpler numbering system to represent curvature which may be modeled by mapping the radius of curvature values to a discrete set of binned corner numbers. The [`base::cut()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/cut) *R* function allows us to perform this mapping using custom bin widths and ordered bin labels.

In the following example, we use the 10 point *turn severity* scale and associated radius of curvature values suggested in [*To see the invisible ( (Volume 1)*](http://therallydriver.com/index.php/en/to-see-the-invisible); low values represent smaller radius of curvature values and hence tighter corners.  Note that we further are labeling straights as category 11 curves.

```{r}
invisible_bins = c(0, 10, 15, 20, 27.5, 35,
                    45, 60, 77.5, 100, 175, Inf)

route_convexity$invisible_ts = cut(route_convexity$radius,
                                   breaks = invisible_bins,
                                   labels = 1:(length(invisible_bins)-1),
                                   ordered_result=TRUE)
```

Let's preview the values:

```{r}
head(route_convexity[,c('ConvexityIndex',
                        'radius', 'invisible_ts')])
```

We can visualise the turn severity values at each 10m sample point along the route:

```{r radius-example}
radius_filter = route_convexity$radius>0 & route_convexity$radius<100

ggplot() + geom_sf(data=utm_routes[1,], color='grey') +
  geom_point(data=route_convexity[radius_filter,],
             
            aes(x=Midpoint_X, y=Midpoint_Y,
                color=invisible_ts), size=0.5)
```

### Percentage in Turn

A really crude metric of how severe the amount of turning is over the duration of a stage might be indicated by the root mean square of the curvature:

```{r}
rms = function(x) sqrt(mean(x^2))

rms(route_convexity$ConvexityIndex)
```

Another possible measure would be the percentage of time in turn, relative to a minimum amount of curvature (or maximum radius of curvature).

For example, we can count the number of segments for which the convexity index is above a particular threshold:

```{r}
pc_in_turn_convexity = function(route, min_convexity=0.1) {
   sum(route$ConvexityIndex > min_convexity) / nrow(route)
}

pc_in_turn_convexity(route_convexity)
```

Alternatively, we might calculate the percentage of segments for which the radius of curvature is less than some maximum value:

```{r}
pc_in_turn_curvature = function(route, max_curvature=80) {
   sum(route$radius < max_curvature) / nrow(route)
}

pc_in_turn_curvature(route_convexity)
```

Another more segmented profile could give the percentage time per turn severity:

```{r}
turn_severity = route_convexity %>%
                  group_by(invisible_ts) %>% 
                  summarise(n = n())

turn_severity$n  = turn_severity$n / nrow(route_convexity)
```

For the non-straight sections, we can plot their distribution:

```{r pc_in_turn_severity_bar}
ggplot(turn_severity[turn_severity$invisible_ts < 11,]) +
  geom_bar(aes(x=invisible_ts, y=n),
           stat='identity')
```

We can also compare the percentage of segments spent on the straight against the percentage involved in some sort of turn: 

```{r pc_in_turn}
pc_in_turn = route_convexity %>%
                  # Create a logical group
                  group_by(in_turn = invisible_ts<11) %>% 
                  summarise(n = n())

pc_in_turn$n  = pc_in_turn$n / nrow(route_convexity)

ggplot(pc_in_turn) +
  geom_bar(aes(x=in_turn, y=n),
           stat='identity')
```


## Estimating Speeds

As well as providing an estimate of turn severity and identifying straights, we can also use the radius of curvature, or the turn severity index, as a crude estimate of maximum speed along that segment. Once again volume 1 of the *To see the invisible* pacenotes tutorial provides some example speeds for each turn intensity value, noting that speeds are likely different on sealed tarmac surfaces or unsealed gravel surfaces:

```{r}
# Speeds in km/h
invisible_speeds = c(10, 40, 50, 60, 70, 80,
                     95, 110, 120, 130, 145)
```

We can use these speeds to model (very crudely!) expected travel times along different parts of the route, or at least provide a crude baseline estimate, as well as a crude estimate over the overall average route speed, which is often regulated under rally competition regulations.

```{r}
route_convexity$invisible_sp = cut(route_convexity$radius,
                                   breaks = invisible_bins,
                                   labels = invisible_speeds,
                                   ordered_result=TRUE)

# Cast speed as factor, via character, to integer
route_convexity$invisible_sp = as.integer(as.character(route_convexity$invisible_sp))

head(route_convexity[,c('ConvexityIndex', 'radius',
                        'invisible_ts', 'invisible_sp')])
```

From the speed estimates, can we then perform a crude time estimate for the stage time, assuming infinite acceleration and deceleration? (This may turn out to be not very realistic at all, of course!)

For each segment, identify the segment transit time as the length of the segment divide by the speed:

```{r}
route_convexity$invisible_time = 10/ (route_convexity$invisible_sp * 1000 /3600)

duration = sum(route_convexity$invisible_time) 
paste0(duration %/% 60, 'm', round(mod(duration, 60),1), 's' )
#11m5s
# Actual stage time: 12:05.7	
# via https://www.ewrc-results.com/results/41079-rallye-automobile-de-monte-carlo-2021/?s=293051
# Difference: (11*60+5) / (12 * 60 + 5)  = 0.917 so ~9% error?
```

A better approach would be to try to model acceleration and evolve the speed across each segment.

A crude acceleration model might look to accelerate up into a faster segment, and decelerate down into a slower one:

```{r}
sp = route_convexity$invisible_sp
sp[1] = 30 # nominal starting speed

# Crude acceleration / brake weights
acc = 1
dec = 1
for (i in 2:(length(sp)-1)) {
  # Accelerate up
  if (sp[i-1]<=sp[i]) sp[i] = (sp[i-1] + acc * sp[i]) / (1+acc)
                                  
  # Decelerate down
  if (sp[i]>sp[i+1]) sp[i] = (dec * sp[i] + sp[i+1]) / (1+dec)
}

route_convexity$acc_sp = sp
route_convexity$acc_sp[length(sp)] = route_convexity$invisible_sp[length(sp)]

head(route_convexity[,c('invisible_sp','acc_sp')])
```

And the new time model?

```{r}
route_convexity$invisible_time2 = 10/ (route_convexity$acc_sp * 1000 /3600)

duration2 = sum(route_convexity$invisible_time2) 
paste0(duration2 %/% 60, 'm', round(mod(duration2, 60),1), 's' )
# 12m9.1s
# Difference: (12*60+9) / (12 * 60 + 5)  = 1.006 so minimal % error?
```