```{r cache = T, echo = F, message=F}
# 1Allow knitr to continue to execute even in the presence of errors
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(fig.path = "images/stage-route-preview-")
```

# Working with Route Data

Having learned how to load in the route data in the previous chapter, we'll use this chapter to explore in a little bit more detail what we can actually do with it.

## Loading in the Route Data

Each chapter in this book is intended to be self-contained in terms of the code contained within the chapter. So we'll need to start by loading the data in.

For convenience, I'll load the data in from the geojson file that we created from the original KML data file in the previous chapter.

```{r warning=FALSE, message=FALSE}
library(sf)

# We previously converted the KML file to a geojson file
geojson_filename = 'montecarlo_2021.geojson'

# The geojson file contains essentially the same data (except the zeroed Z values)
# as the original KML file
geojson_sf = st_read(geojson_filename)
```

Howsoever we have loaded in the data, we should have some sort of data structure that contains one or more routes. In the case of our loaded in KML file, or the geojson object,  each route is presented in the form of a linestring, made up as a list of consecutive co-ordinate pairs along the route.

To simplify the view of the data we might expect to see, let's cast the original KML object obtained from the WRC KML file directly into a dataframe:

```{r}
kml_df = as.data.frame(geojson_sf)
kml_df
```

## Access a specific stage

We can access a specific stage, which is to say, a specific row of the features dataframe, by direct indexing:

```{r}
geojson_sf[3,]
```

We can find the length of each stage route using the `sf::st_length()` function:

```{r}
st_length(geojson_sf)
```

Or the length of a specific stage:

```{r}
st_length(geojson_sf[1,])
# 21069.62
```

Note that you may need to take this distances with a pinch of salt. The WRC website gives the stage length of SS1 as 20.58 km, for example, not 21.07 km, so it's not clear whether the route in the KML file is the competitive route, or whether is starts and finishes with time controls at either end of the route. It may be that the stage distance also follows the racing line, which is likely to be shorter than than the piecewise linear route we access from the KML file.

There are also several different ways in which we can calculate the distance each of which may have slightly different parameter settings baked in, such as the distance round the equator, that are likely to influence the result of any distance calculations we perform.

It is also worth noting that elevation changes mean that the actual route length is likely to be *greater* than the distance measured over a 2 dimensional route in the XY (longitude-latitude) plane.


## Visual Preview of Route Geodata

Let's generate a quick preview of all the stages. The dataframe contains a column of encoded linestrings. We can preview these directly with a simple plot:

```{r kml_peview}
plot(st_geometry(geojson_sf))
```

This shows the rough outline of each stage, and their relative positions, but represents little more than a back of a napkin sketch.  

## Previews using `ggplot2`

In some circumstances, we may want to work with the route data in `ggplot2`.

The `ggplot2` package is aware of simple features (`sf`) representations, although we need to inform it that such a dataset has been provided via the [`ggplot2::geom_sf()` function](https://ggplot2.tidyverse.org/reference/ggsf.html).

We can preview the data as follows, firstly over all features:

```{r gg-preview}
# https://ggplot2.tidyverse.org/reference/ggsf.html
library(ggplot2)

ggplot(data=geojson_sf) + geom_sf()
```

Note how the co-ordinate axes are rendered in degrees longitude (x coordinate) and degrees latitude (y co-ordinate) using a familiar projection (that is, the x and y axes are in a proportion that you are likely to recognise from most atlases and (web) maps).

The `ggpubr::ggarrange()` function allows us to create an arrangement of plots made up from a list separate plot for each stage. We can also take the opportunity to demonstrate how we might start styling the x-tick labels using the `axis.text.x` parameter to the `ggplot::theme()`:

```{r gg-pubr}
stageplot = function(x, output) {
  ggplot(data=x) + geom_sf() +
    theme(axis.text.x = element_text(angle = 90,
                                     face='italic',  size = 7),
          axis.text.y = element_text(face='italic',  size = 7))
}
 

stages_kml = kml_df$geometry

#apply(X, MARGIN, FUN, …)
ggpubr::ggarrange(plotlist=lapply(stages_kml, stageplot))
```

We can visualise the data for just a single route by limiting the data we pass to the `ggplot()` function:

```{r gg-single-route}
g = ggplot(data=geojson_sf[1,]) + geom_sf()
g
```
In rendering this map, note how we can assign the plot to a variable (`g`) and then display the map by referencing the variable.

## Identifying the Start and End of the Stage

We can grab the co-ordinates of locations at the start and end of the stage, assuming that the linestrings are presented in the order *start-finish*:

```{r}
stage_coords = as.data.frame(st_coordinates(kml_df[1, ]$geometry))

stage_start_coords = stage_coords[1, ]
stage_end_coords = stage_coords[nrow(stage_coords), ]

first_and_last = stage_coords[c(1,nrow(stage_coords)),]
first_and_last$label=c('SS1 stage start', 'SS1 stage end')
first_and_last$typ=c('start', 'stop')
first_and_last$color=c('green', 'red')
first_and_last
```

We can then highlight these features on our stage plot:

```{r gg-route-start-stop}
g + 
  geom_point(aes(x=X, y=Y,  color = typ),size =5, alpha = 0.5,
             data=first_and_last) +
  scale_color_manual(values=c("blue", "red"))
```

## Extracting Route Fragments

We can readily extract a fragment from a line by indexing against the line segment indices. For example, to gran the first 10 straighline segments from the linestring we can extract just those coordinates and build a new linestring from them:

```{r}
cc = as.data.frame(st_coordinates(geojson_sf[1,]))[1:10,]

route_segment = st_linestring(cbind(x=cc[,1], y=cc[,2]))

# We can plot the linestring using the same sort of geo projection
# as was applied to the original route
ggplot(route_segment) + geom_sf(crs=st_crs(geojson_sf[1,]))

```

## Previews Using `leaflet` Interactive Maps

Simple plots of the route gives us a snapshot of the shapes of various routes and their relative positioning, but it's far more useful to preview the routes over a simple map background, such as an interactive OpenStreetMap map.

The [`leaflet` package](https://rstudio.github.io/leaflet/) provides a range of tools for working with the *leaflet* Javascript package to visualise such maps.

The `leaflet::leaflet` function creates interactive maps based around map tiles that can be retrieved from a variety of map tile providers. Some providers may require an API key or access credentials. For moderate use, we can freely access *OpenStreetMap* or [*Stamen*](http://maps.stamen.com/) tiles in a variety of themes without any form of authentication:

```{r warning=FALSE, message=FALSE}
library(leaflet)

# Alternatively, we might set:
# stages_kml = kml[,'geometry']

# Note that leaflet seems to requires us to to have dropped 
# the Z and M dimensions from the linestrings
leaflet(stages_kml) %>% 
  addProviderTiles("Stamen.TonerLite",
                   # Other themes include:
                   # Stamen.TonerLite, Stamen.Terrain, OpenStreetMap.Mapnik
                   # Stamen.TerrainBackground, OpenTopoMap
                   group = "OSM") %>% 
  addPolylines(color = "red", weight = 3)
```

The *Stamen.TonerLite* theme is very clear for identifying where the stages are whilst the *OpenTopoMap* and *Stamen.TerrainBackground* themes are using for showing relief, particularly in mountainous or hilly stages.

As well as previewing all the stage routes in the features collection, we can extract the linestring geometry for a single route:

```{r}
kml_geom = kml_df[1, ]$geometry
kml_geom
# kml_df[kml_df$Name=='SS 1' , ] # Get row
```

We can now generate a quick map preview of that particular stage:

```{r}
leaflet(stages_kml[1]) %>% 
  addProviderTiles("OpenTopoMap", group = "OSM") %>% 
  addPolylines(color = "red", weight = 5)
```

We can also highlight the stage start and end points as simple circle markers, as pin "balloon" markers, in this case with a fixed label to annotate the stage start and finish:

```{r}
# Create a palette that maps factor levels to colors
pal <- colorFactor(c("darkgreen", "darkred"),
                   domain = c("start", "stop"))

m = leaflet(stages_kml[1]) %>% 
      addProviderTiles("OpenTopoMap", group = "OSM") %>% 
      addPolylines(color = "red", weight = 5)

m %>%
  addMarkers(~X, ~Y, #popup = ~label,
             label = ~label,
             labelOptions = labelOptions(noHide = T,
                                         direction = "bottom",
                                         style = list(
                                          "color" = 'black',
                                          "font-family" = "serif",
                                          "font-style" = "italic",
                                          "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
                                          "font-size" = "12px",
                                          "border-color" = "rgba(0,0,0,0.5)"
                                          )),
            data=first_and_last) %>%
  addCircleMarkers(~X, ~Y,  radius =10, color = ~pal(typ),
                   stroke = FALSE, fillOpacity = 0.5,
                   data=first_and_last)
```

As well as using default markers, we can define custom markers:

```{r}
# https://fontawesome.com/v4.7.0/icons/ 
start_stop_markers <- awesomeIconList(
  "start" = makeAwesomeIcon(
    icon = "play-circle",
    markerColor = "blue",
    library = "fa"
  ),
    "stop" = makeAwesomeIcon(
    icon = "stop-circle",
    markerColor = "red",
    library = "fa"
  )
)
```

We can add the markers directly to our earlier created leaflet map object:

```{r}
m %>% 
  addAwesomeMarkers(~X, ~Y,
                    icon = ~ start_stop_markers[typ],
                    label = ~ label,
                    data=first_and_last)
m
```

We can save a map to an image file using the `mapview` package:

```{r}
library(mapview)
# mapview also provides tools for generating leaflet maps from R
image_fn = "demomap.png"

mapshot(m, file = image_fn)
```

We can then review the saved image captured from the map:

```{r embed-demo-map}
knitr::include_graphics(image_fn)
```

## Creating Buffer regions around features

One thing we note from the map is that the display is cropped quite close to the extent of the displayed stage. It would be useful to provide a small buffer around the stage to provide a margin that allows us to see the start and stop clearly.

We can get the bounds of the stage quite simply:

```{r}
stage_bbox = st_bbox(stages_kml[1])
stage_bbox
```

A convenient way of representing the bounding box is to use the *extent()* function from the `raster` package to create an `Extent` class object:

```{r}
raster::extent(stage_bbox)
```

In some functions, we may be able to pass an *extent* object in directly as a function argument.

We can also create a dataframe containing those limits:

```{r}
ex.df <- data.frame(x= c(stage_bbox[['xmin']], stage_bbox[['xmax']]), 
                    y= c(stage_bbox[['ymin']], stage_bbox[['ymax']]))
ex.df
```

The contents of the dataframe representation can be cast as a spatial points object containing two features, whose extent matches the original extent:

```{r}
# For SpatialPoints
library(sp)

bb_sp = SpatialPoints(ex.df)
bb_sp
```

This is useful because we can  create a "buffer" region around those points that will "push out" the extent and give us some co-ordinates that represent a buffered area around the original path.

Let's add a buffer around the trace of five hundred meters:

```{r}
library(spatialEco)

#In passing, recall that WGS84 (latlon) is st_crs(4326) 
proj4string(bb_sp) <- '+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs' 

# Create a new, buffered region
stage_bbox = st_bbox(geo.buffer(x=bb_sp, r=500))

# Also update the dataframe of points with these bounds
ex.df <- data.frame(x= c(stage_bbox[['xmin']], stage_bbox[['xmax']]), 
                    y= c(stage_bbox[['ymin']], stage_bbox[['ymax']]))

stage_bbox
```

We can pass these new bounds into the leaflet widget to try to encourage it to show a little more of the map, although noting that it doesn't always force the issue if the buffer isn't big enough...

```{r}
leaflet(stages_kml[1]) %>% 
  fitBounds(stage_bbox[['xmin']], stage_bbox[['ymin']],
            stage_bbox[['xmax']], stage_bbox[['ymax']])  %>% 
  #fitBounds(5.879966, 44.732916,
  #          5.944508, 44.822432)  %>% 
  addProviderTiles("OpenTopoMap", group = "OSM") %>% 
  addPolylines(color = "red", weight = 5)
```

## Adding Additional Features to `leaflet` maps

Several packages exist that allow us to add additional controls and features to interactive *leaflet* maps, including "minimap" navigational support, the ability to add logos and charts to maps, and support for arbitrary HTML and chart content within popups.

### Add "minimap" navigation

To support navigation(?!), we can also add a "minimap", with it's own location and display settings, as well as its own tileset, to show more clearly where in the world is being displayed by the current view.

```{r}
m %>%
 addMiniMap( tiles="Stamen.TonerLite", toggleDisplay = TRUE,
    position = "bottomleft")
```

### Add Custom Symbol Legends

We could add eg rally signs to start and end of stage (and if we have richer detail, radio points, etc). The [`leaflegend`](https://roh.engineering/post/2021-02-04-introduction-to-leaflegend/) R package provides several utilities for creating custom icons.

Rally stage symbols can be found in various custom fonts such as [*dafont/rally-symbols*](https://www.dafont.com/rally-symbols.font) or [*wfonts/rally-symbols*](https://www.wfonts.com/font/rally-symbols).

#### Adding a Logo
The [`r-spatial::leafem` package](https://github.com/r-spatial/leafem) adds several extension features to leaflet maps. For example, we can add a logo to the map:

```{r}
img <- "https://acm.mc/wp-content/uploads/2020/04/Plaque-rouge-2021.png"

# Can we control the size? Or do we have to make a square logo?
m %>% leafem::addLogo(img)
```


### Adding Mini-Charts to Maps

The [`leaflet.minicharts`](https://cran.r-project.org/web/packages/leaflet.minicharts/vignettes/introduction.html) R package allows simple charts to be displayed at specific locations on a map.

### Adding HTML content to popups

The [`leafpop`](https://github.com/r-spatial/leafpop) R package allows us to add HTML tables, images or graphs to `leaflet` or `mapview` popups, as well as `ggplot2` graphics and `htmlwidgets`.

This packages might be useful for adding stage results displays to maps, for example.

## Using `mapview` To Create *leaflet* Maps

As well as using the `leaflet` package to generate *leaflet* maps, the [`mapview` R package](https://r-spatial.github.io/mapview/) provides an alternative way of .

```{r message=FALSE}
library(mapview)

mv = mapview(geojson_sf)
mv
```

The `r-spatial/leafem` package that we can use to enrich maps using the `leaflet` package was in part developed as an abstraction of functions contained within the `mapview` package; both `mapview` and `leafem` are maintained by the [*r-spatial* Github organisation](https://github.com/r-spatial).
