```{r cache = T, echo = F, message=F}
# 15Allow knitr to continue to execute even in the presence of errors
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(fig.path = "images/terrain-features-raster-")
```
# Determining Terrain Features from Raster Data

Based on the elevation data associated with a raster image, we can depict the raster image in various ways. For example, we can select different colour plots to colour the tiles, as well as adding contour lines.

## Load in Base Data 

As ever, let's load in our stage data:

```{r}
geojson_filename = 'montecarlo_2021.geojson'
geojson_sf = sf::st_read(geojson_filename)
```

And also load in the elevation data raster:

```{r message=FALSE, warning=FALSE}
library(raster)
library(rayshader)

# Previously downloaded TIF digital elevation model (DEM) file
stage_tif = "stage_elevation.tif"

# Load in the previously saved image raster
elev_img = raster(stage_tif)
auto_zscale = geoviz::raster_zscale(elev_img)

# Note we can pass in a file name or a raster object
elmat = raster_to_matrix(stage_tif)
```

## Rendering Terrain Over Raster Images

By default, a plot of a raster image will render a colour mapping sensitive to elevation:

```{r elev-preview}
plot(elev_img, col = terrain.colors(25), axes=FALSE)
```

As well as height / elevation information, we can also extract additional information from the raster image using the `raster::terrain()` function. This function can compute various terrain characteristics from a raster with elevation data, ideally in a projected (planar) raster data image with units in meters.

Terrain features include *aspect*, *slope* and *flowdir* (the direction of water flow as determined by the direction in which the greatest drop in elevation (or smallest rise) is to be found at any given point). *TPI* (Topographic Position Index, *"the difference between the value of a cell and the mean value of its 8 surrounding cells"), *TRI* (Terrain Ruggedness Index, *"the mean of the absolute differences between the value of a cell and the value of its 8 surrounding cells"*) and *roughness* (*"the difference in elevation between the maximum and the minimum value of a cell and its 8 surrounding cells"*) terrain measures are also supported:

```{r terrain-preview}
plot(terrain(elev_img, "TRI"),
     # Remove legends, bounding box etc
      bty="n", box=FALSE, axes=F, frame.plot=F, legend=F)
```

To give the appearance of relief as highlighted by shadow, we can shade the terrain using the the `raster::hillshade()` function with slope and aspect values determined by the `raster::`terrain()` function:

```{r hill-preview}
hill_shadows = hillShade(slope = terrain(elev_img, "slope"),
                         aspect = terrain(elev_img, "aspect"),
                         # azimuthal direction of sun in degrees:
                         angle = 45, 
                         add=TRUE)
plot(hill_shadows)
```


## Adding Contours to `raster` Images

Consider again our original raster image, plotted here using a "terrain style" colour palette:

```{r terrain-preview2}
plot(elev_img, col = terrain.colors(25), axes=FALSE)
```

We can use the `raster::rasterToContour` function to generate a set of contour lines over the raster:

```{r terrain-contour}
# Plot the original raster image
plot(elev_img, col = terrain.colors(25), axes=FALSE)

# Add a contour plot over the raster image
plot (rasterToContour(elev_img), add=TRUE)

#nlevels - number of contour levels desired iff levels is not supplied.
#levels - numeric vector of levels at which to draw contour lines.
```

We have a certain amount of control over the contours that are plotted, notably through the following two parameters that can be passed through the `rasterToContour()` function to the `grDevices::contourLines` function [[docs](https://www.rdocumentation.org/packages/grDevices/versions/3.0.3/topics/contourLines)] that actually generates the contour lines:

- `nlevels`: the number of contour levels desired iff `levels` is not supplied;
- `levels`: a numeric vector of levels at which to draw the contour lines.

So for example, let's set the contour lines to appear at elevations of 1000m and 1500m:

```{r terrain-raster-contour}
# Plot the original raster image
plot(elev_img, axes=FALSE)

# Add a contour plot over the raster image
plot(rasterToContour(elev_img, levels=c(1000, 1500)), add=TRUE)
```
Alternatively, we may choose to let "the algorithm" decide the levels whilst we still determine how many contour levels there are to be:

```{r terrain-raster-contour2}
# Plot the original raster image
plot(elev_img, axes=FALSE)

# Add a contour plot over the raster image
plot (rasterToContour(elev_img, nlevels=4),
      add=TRUE)
```

Another `raster` package function, `raster::contour`, produces a contour plot with labeled contour values:

```{r contour-plot}
contour(elev_img)
```

Again, we can overplot this on the original raster image, optionally setting the contour line color:

```{r contour-plot2}
plot(elev_img)

contour(elev_img, add=TRUE, col='navyblue')
```

